---
title: "Shiny"
author: "Eric Archer"
date: "3/5/2018"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  comment = "",
  prompt = TRUE, 
  fig.align = "left"
)

rm(list = ls())
library(shiny)
library(tidyverse)
```

Shiny tutorial: http://shiny.rstudio.com/tutorial/

The basics
```{r}
# setup the HTML user interface
ui <- fluidPage()

# define a function containing code being executed in UI
server <- function(input, output) {}

# run the app
shinyApp(ui = ui, server = server)
```


Simple app
```{r}
ui <- fluidPage(
  numericInput(inputId = "n", "Sample size", value = 25),
  plotOutput(outputId = "hist")
)

server <- function(input, output) {
  output$hist <- renderPlot({
    hist(rnorm(input$n))
  })
}

shinyApp(ui = ui, server = server)
```

Inputs and Outputs
```{r}
ui <- fluidPage(
  numericInput("numInput", "numericInput", 5),
  textOutput("numOutput"),
  sliderInput("slInput", "sliderInput", 100, 200, 150),
  textOutput("slOutput"),
  checkboxGroupInput("chgrpInput", "checkboxGroupInput", list("a" = 1, "b" = 2, "c" = 3)),
  textOutput("chgrpOut")
)

server <- function(input, output) {
  output$numOutput <- renderText(print(input$numInput))
  output$slOutput <- renderText(print(input$slInput))
  output$chgrpOut <- renderText(print(paste(input$chgrpInput, collapse = " / ")))
}

shinyApp(ui = ui, server = server)
```


Reactivity

render*()
```{r}
ui <- fluidPage(
  sliderInput("n", "Number of samples", 10, 1000, 100, 10),
  sliderInput("slope", "Slope", -1, 1, 0, 0.1),
  sliderInput("intercept", "Intercept", -1, 1, 0, 0.1),
  numericInput("sd", "SD", 0.1, min = 0, step = 0.05),
  plotOutput("plot"),
  textOutput("smryTO"),
  verbatimTextOutput("smryVTO")
)

server <- function(input, output) {
  output$plot <- renderPlot({
    x <- runif(input$n)
    y <- input$intercept + input$slope * x + rnorm(input$n, 0, input$sd)
    plot(x, y, ylim = c(-1, 1))
  })
  output$smryTO <- renderPrint({ summary(runif(input$n)) })
  output$smryVTO <- renderPrint({ summary(runif(input$n)) })
}

shinyApp(ui = ui, server = server)
```

reactive({ code with reactive values that returns object}) : creates a "function" that saves validated reactive values
```{r}
ui <- fluidPage(
  sliderInput("n", "Number of samples", 10, 1000, 100, 10),
  sliderInput("slope", "Slope", -1, 1, 0, 0.1),
  sliderInput("intercept", "Intercept", -1, 1, 0, 0.1),
  numericInput("sd", "SD", 0.1, min = 0, step = 0.05),
  plotOutput("plot"),
  h4("x Summary"),
  verbatimTextOutput("smryX"),
  h4("y Summary"),
  verbatimTextOutput("smryY")
)

server <- function(input, output) {
  pts <- reactive({
    x <- runif(input$n)
    y <- input$intercept + input$slope * x + rnorm(input$n, 0, input$sd)
    list(x = x, y = y)
  })
  
  output$plot <- renderPlot({
    plot(pts()$x, pts()$y, ylim = c(-1, 1))
  })
  output$smryX <- renderPrint({ summary(pts()$x) })
  output$smryY <- renderPrint({ summary(pts()$y) })
}

shinyApp(ui = ui, server = server)
```


isolate({ code with reactiveValues that returns object }) : creates a non-reactive value
```{r}
ui <- fluidPage(
  sliderInput("n", "Number of samples", 10, 1000, 100, 10),
  sliderInput("slope", "Slope", -1, 1, 0, 0.1),
  sliderInput("intercept", "Intercept", -1, 1, 0, 0.1),
  numericInput("sd", "SD", 0.1, min = 0, step = 0.05),
  plotOutput("plot"),
  h4("x Summary"),
  verbatimTextOutput("smryX")
)

server <- function(input, output) {
  x <- reactive({runif(input$n)})
  
  output$plot <- renderPlot({
    y <- isolate({
      input$intercept + input$slope * x() + rnorm(length(x()), 0, input$sd)
    })
    plot(x(), y, ylim = c(-1, 1))
  })
  output$smryX <- renderPrint({ summary(x()) })
}

shinyApp(ui = ui, server = server)
```

observeEvent(reactiveValues, { code }) : execute code when specified reactive values are changed
```{r}
ui <- fluidPage(
  sliderInput("n", "Number of samples", 10, 1000, 100, 10),
  sliderInput("slope", "Slope", -1, 1, 0, 0.1),
  sliderInput("intercept", "Intercept", -1, 1, 0, 0.1),
  numericInput("sd", "SD", 0.1, min = 0, step = 0.05),
  plotOutput("plot"),
  actionButton("estimate", "Estimate regression"),
  verbatimTextOutput("smry")
)

server <- function(input, output) {
  pts <- reactive({
    x <- runif(input$n)
    y <- input$intercept + input$slope * x + rnorm(length(x), 0, input$sd)
    data.frame(x = x, y = y)
  })
  
  output$plot <- renderPlot({ plot(pts(), ylim = c(-1, 1)) })
  
  observeEvent(input$estimate, {
    pt.lm <- lm(y ~ x, pts())
    output$plot <- renderPlot({
      plot(pts(), ylim = c(-1, 1)) 
      abline(coef(pt.lm), color = "red")
    })
    output$smry <- renderPrint({ summary(pt.lm) })
  })
}

shinyApp(ui = ui, server = server)
```

observe({ code with reactive values }) : executes all code when reactive values are changed. like reactive({}), but does not create an object that can be referenced later.
```{r}
ui <- fluidPage(
  sliderInput("n", "Number of samples", 10, 1000, 100, 10),
  sliderInput("slope", "Slope", -1, 1, 0, 0.1),
  sliderInput("intercept", "Intercept", -1, 1, 0, 0.1),
  numericInput("sd", "SD", 0.1, min = 0, step = 0.05),
  plotOutput("plot"),
  verbatimTextOutput("smry")
)

server <- function(input, output) {
  pts <- reactive({
    x <- runif(input$n)
    y <- input$intercept + input$slope * x + rnorm(length(x), 0, input$sd)
    data.frame(x = x, y = y)
  })
  
  observe({
    if(input$n > 500) { # this if statement must be in a reactive function because it has a reactive value
      output$smry <- renderText({ cat("") })
      output$plot <- renderPlot({ plot(pts(), ylim = c(-1, 1)) })
    } else {
      pt.lm <- lm(y ~ x, pts())
      output$smry <- renderPrint({ summary(pt.lm) })
      output$plot <- renderPlot({
        plot(pts(), ylim = c(-1, 1))
        abline(coef(pt.lm), color = "red")
      })
    }
  })
}

shinyApp(ui = ui, server = server)
```

eventReactive(reactive values, { code }) : returns value generated by code when reactive values are changed.
```{r}
ui <- fluidPage(
  sliderInput("n", "Number of samples", 10, 1000, 100, 10),
  sliderInput("slope", "Slope", -1, 1, 0, 0.1),
  sliderInput("intercept", "Intercept", -1, 1, 0, 0.1),
  numericInput("sd", "SD", 0.1, min = 0, step = 0.05),
  actionButton("go", "Go"),
  plotOutput("plot"),
  verbatimTextOutput("smry")
)

server <- function(input, output) {
  pts <- eventReactive(input$go, {
    df = data.frame(x = runif(input$n))
    df$y <- input$intercept + input$slope * df$x + rnorm(nrow(df), 0, input$sd)
    list(df = df, lm = lm(y ~ x, df))
  })
  
  output$plot <- renderPlot({
    plot(pts()$df, ylim = c(-1, 1)) 
    abline(coef(pts()$lm), color = "red")
  })
  output$smry <- renderPrint({ summary(pts()$lm) })
}

shinyApp(ui = ui, server = server)
```


reactiveValues(elements) : create list of elements that are reactive values and can be changed programmatically.
```{r}
ui <- fluidPage(
  sliderInput("n", "Number of samples", 10, 1000, 100, 10),
  sliderInput("intercept", "Intercept", -1, 1, 0, 0.1),
  sliderInput("slope", "Slope", -1, 1, 0, 0.1),
  numericInput("sd", "SD", 0.1, min = 0, step = 0.05),
  plotOutput("plot"),
  verbatimTextOutput("smry"),
  sliderInput("intercept.test", "Test - Intercept", -1, 1, 0, 0.1),
  sliderInput("slope.test", "Test - Slope", -1, 1, 0, 0.1)
)

server <- function(input, output) {
  line.fit <- reactiveValues(intercept = NA, slope = NA)
  
  pts <- reactive({
    df <- data.frame(x = runif(input$n))
    df$y <- input$intercept + input$slope * df$x + rnorm(nrow(df), 0, input$sd)
    lm.fit <- lm(y ~ x, df)
    line.fit$intercept <- coef(lm.fit)[1]
    line.fit$slope <- coef(lm.fit)[2]
    list(df = df, lm = lm.fit)
  })
  
  observeEvent(input$intercept.test, ({ line.fit$intercept <- input$intercept.test }))
  observeEvent(input$slope.test, ({ line.fit$slope <- input$slope.test }))
  
  output$plot <- renderPlot({
    plot(pts()$df, ylim = c(-1, 1)) 
    abline(a = line.fit$intercept, b = line.fit$slope, color = "red")
  })
  output$smry <- renderPrint({ summary(pts()$lm) })
}

shinyApp(ui = ui, server = server)
```


Layout
```{r}
ui <- fluidPage(
  sidebarLayout(
    sidebarPanel(
      sliderInput("n", "Number of samples", 10, 1000, 100, 10),
      sliderInput("intercept", "Intercept", -1, 1, 0, 0.1),
      sliderInput("slope", "Slope", -1, 1, 0, 0.1),
      numericInput("sd", "SD", 0.1, min = 0, step = 0.05)
    ),
    mainPanel(
      plotOutput("plot"),
      verbatimTextOutput("smry"),
      sliderInput("intercept.test", "Test - Intercept", -1, 1, 0, 0.1),
      sliderInput("slope.test", "Test - Slope", -1, 1, 0, 0.1)
    )
  )
)

server <- function(input, output) {
  line.fit <- reactiveValues(intercept = NA, slope = NA)
  
  pts <- reactive({
    df <- data.frame(x = runif(input$n))
    df$y <- input$intercept + input$slope * df$x + rnorm(nrow(df), 0, input$sd)
    lm.fit <- lm(y ~ x, df)
    line.fit$intercept <- coef(lm.fit)[1]
    line.fit$slope <- coef(lm.fit)[2]
    list(df = df, lm = lm.fit)
  })
  
  observeEvent(input$intercept.test, ({ line.fit$intercept <- input$intercept.test }))
  observeEvent(input$slope.test, ({ line.fit$slope <- input$slope.test }))
  
  output$plot <- renderPlot({
    plot(pts()$df, ylim = c(-1, 1)) 
    abline(a = line.fit$intercept, b = line.fit$slope, color = "red")
  })
  output$smry <- renderPrint({ summary(pts()$lm) })
}

shinyApp(ui = ui, server = server)
```


```{r}
ui <- fluidPage(
  sidebarLayout(
    sidebarPanel(
      sliderInput("n", "Number of samples", 10, 1000, 100, 10),
      sliderInput("intercept", "Intercept", -1, 1, 0, 0.1),
      sliderInput("slope", "Slope", -1, 1, 0, 0.1),
      numericInput("sd", "SD", 0.1, min = 0, step = 0.05)
    ),
    mainPanel(
      tabsetPanel(
        tabPanel(
          "Plot",
          wellPanel(
            plotOutput("plot"),
            sliderInput("intercept.test", "Test - Intercept", -1, 1, 0, 0.1),
            sliderInput("slope.test", "Test - Slope", -1, 1, 0, 0.1)
          )
        ),
        tabPanel("Fit", verbatimTextOutput("smry"))
      )
    )
  )
)

server <- function(input, output) {
  line.fit <- reactiveValues(intercept = NA, slope = NA)
  
  pts <- reactive({
    df <- data.frame(x = runif(input$n))
    df$y <- input$intercept + input$slope * df$x + rnorm(nrow(df), 0, input$sd)
    lm.fit <- lm(y ~ x, df)
    line.fit$intercept <- coef(lm.fit)[1]
    line.fit$slope <- coef(lm.fit)[2]
    list(df = df, lm = lm.fit)
  })
  
  observeEvent(input$intercept.test, ({ line.fit$intercept <- input$intercept.test }))
  observeEvent(input$slope.test, ({ line.fit$slope <- input$slope.test }))
  
  output$plot <- renderPlot({
    plot(pts()$df, ylim = c(-1, 1)) 
    abline(a = line.fit$intercept, b = line.fit$slope, color = "red")
  })
  output$smry <- renderPrint({ summary(pts()$lm) })
}

shinyApp(ui = ui, server = server)
```

```{r}
ui <- fluidPage(
  sidebarLayout(
    sidebarPanel(
      h4("Simulated Data"),
      sliderInput("n", "Number of samples", 10, 300, 100, 10),
      sliderInput("intercept", "Intercept", -1, 1, 0, 0.01),
      sliderInput("slope", "Slope", -1, 1, 0, 0.01),
      sliderInput("sd", "SD", 0, 1, 0.1, 0.01)
    ),
    mainPanel(
      tabsetPanel(
        tabPanel(
          "Plot",
          wellPanel(
            plotOutput("plot"),
            h5("Test fit"),
            sliderInput("intercept.test", "Intercept", -1, 1, 0, 0.01),
            sliderInput("slope.test", "Slope", -1, 1, 0, 0.01),
            h5("Residuals"),
            verbatimTextOutput("residuals"),
            h5("Sums of squares"),
            textOutput("ss")
          )
        ),
        tabPanel("Fit", verbatimTextOutput("smry"))
      )
    )
  )
)

server <- function(input, output) {
  line.fit <- reactiveValues(intercept = 0, slope = 0)
  
  pts <- reactive({
    df <- data.frame(x = runif(input$n))
    df$y <- input$intercept + input$slope * df$x + rnorm(nrow(df), 0, input$sd)
    pts.lm <- lm(y ~ x, df)
    line.fit$intercept <- coef(pts.lm)[1]
    line.fit$slope <- coef(pts.lm)[2]
    list(df = df, lm = pts.lm)
  })
  
  observeEvent(input$intercept.test, ({ line.fit$intercept <- input$intercept.test }))
  observeEvent(input$slope.test, ({ line.fit$slope <- input$slope.test }))
  
  observe({
    test.resid <- line.fit$intercept + line.fit$slope * pts()$df$x 
    output$residuals <- renderPrint({ summary(test.resid) })
    output$ss <- renderText({ sum((test.resid - pts()$df$y) ^ 2) })
  })
  
  output$plot <- renderPlot({
    test.resid.y <- line.fit$intercept + line.fit$slope * pts()$df$x
    plot(pts()$df, ylim = c(-1, 1)) 
    abline(a = line.fit$intercept, b = line.fit$slope, col = "red", lwd = 2)
    segments(
      pts()$df$x, pts()$df$y,
      pts()$df$x, test.resid.y,
      col = "blue", lty = "dashed"
    )
  })
  output$smry <- renderPrint({ summary(pts()$lm) })
}

shinyApp(ui = ui, server = server)
```