---
title: "R Markdown"
output: pdf_document
---

**Tutorial**: [http://rmarkdown.rstudio.com/index.html](http://rmarkdown.rstudio.com/index.html)  
**Cheatsheet**: [http://www.rstudio.com/wp-content/uploads/2016/03/rmarkdown-cheatsheet-2.0.pdf](http://www.rstudio.com/wp-content/uploads/2016/03/rmarkdown-cheatsheet-2.0.pdf)  
**Reference Guide**: [https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf](https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf)

***

### Code Chunks

To create a simple code chunk, you can use:

* keyboard shortcuts
     +  Windows: CTL+ALT+i
     +  Mac: Cmd+Option+i
* Insert from toolbar
* type in by hand  

This chunk executes and shows the command:
```{r}
z <- 1
```

The code in this chunk is not visible, but the result is:
```{r, echo = FALSE}
z <- 5
print(z)
```

The code and result are visible from this chunk:
```{r, echo = TRUE}
print(z)
```

Neither the code nor result is visible from this chunk, but the code is still run:
```{r, echo = TRUE, include = FALSE}
x <- rnorm(100, 10, 2)
y <- rnorm(100, 20, 4)
print(x)
# Hello!!!!
```

The plot above shows that this code and the code above was run.
```{r, echo = FALSE}
plot(x, y)
# Hello #2!!!!
```

This code is visible, but is never run:
```{r, echo = TRUE, eval = FALSE}
z <- 250
print(z)
```

We can see that `z` has the same value as when first set:
```{r}
print(z)
```

The comment character can be changed (or eliminated), and a console prompt shown:
```{r, echo = TRUE, comment = "", prompt = TRUE}
a <- 4
b <- 8
# Here is the result
a + b ^ 2
```

Defaults can be changed by setting options to `knitr::opts_chunk$set`:
```{r}
knitr::opts_chunk$set(comment = "*", echo = FALSE)
```

```{r}
# Hello there!!
print(5 * 6)
```


If you are creating a computationally-intensive document, you should cache the results of your chunks:
```{r}
knitr::opts_chunk$set(cache = TRUE)
```

When chunks are cached, it can be good to name them:
```{r random_numbers}
x <- runif(1e5)
y <- mean(x)
cat(y)
```


Full list of knitr chunk options: [https://yihui.name/knitr/options/#other-chunk-options](https://yihui.name/knitr/options/#other-chunk-options)

***

### Inline Code

Code can be placed inline so that results are embedded directly in the text. For instance, a * b ^ 2 = `r a * b ^ 2`. Multiple steps can be done in chunks, then results embedded inline:

```{r}
p <- 0.6
odds.p <- p / (1 - p)
log.odds.p <- log(odds.p)
```
For p = `r p`, the odds(p) = `r odds.p`, and the log-odds(p) = `r round(log.odds.p, 3)`.

***

### Markdown Formatting Syntax

This is a sentence. There is one space following this sentence. The third sentence ends up here.

This is another sentence. If I follow this sentence with two spaces, the third sentence will be a new paragraph.  Like this.

Text can be *italics* or **bold** or ***italics-bold***.

Text can have^superscripts^ and~subscripts~.

See cheatsheet and reference guide for other formatting syntax.

***

### Tables

You can use the vertical separators (i.e., pipe, "|") and horizontal separator (i.e., dash, "-") to create tables. It doesn't have to be neatly aligned as in the examples:

Header 1 | Header 2
-|-
Row1, Col1 | Row1, Col2
Row2, Col1 | Row2, Col2

Here are example default tables for a `data.frame` using `knitr`, `xtable`, and `pander`:
```{r}
df <- mtcars[1:5, ]
library(knitr)
kable(df, caption = "From knitr")
```

```{r, results = "asis"}
library(xtable)
print(xtable(df, caption = "From xtable"), comment = F)
```

```{r}
library(pander)
pander(df, caption = "From pander")
```

***

### Output Formats

See: [http://rmarkdown.rstudio.com/lesson-9.html](http://rmarkdown.rstudio.com/lesson-9.html)

***

### Notebooks

[http://rmarkdown.rstudio.com/r_notebooks.html](http://rmarkdown.rstudio.com/r_notebooks.html)

* Change output format to `html_notebook`.
* When previewed, will create special `.nb.html` file that has controls for hiding code.
* Only one line at a time is sent to console and output from notebook chunks appear beneath the chunk. 